Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-10-02 Dail_Mad_ext.tex
discussion

Our work has highlighted the limitations of classical state-space
models as applied to ecological time series data, and we have
demonstrated how a class of open population $N$-mixture models
proposed by \citet{dail_madsen:2011} overcomes these
limitations. We extended this class of models in several important
ways to accomodate common features of ecological
datasets, including sparse counts, zero-inflation, demographic
stochasticity, and random variation in observation error.
%We made four important developments of the open population
%N-mixture model proposed by \citet{dail_madsen:2011} that make the
%model more applicable to long-term data such as is collected
%by the North American Breeding Bird Survey.
%First, we have
%reconciled the objectives of traditional state-space models
%with open population $N$-mixture models by illustrating how
%classical models of population dynamics can be embedded within
%the framework. Second, we have demonstrated methods of
%accounting for zero-inflation in the time-series. Third, we
%have illustrated how additional random effects such as
%observer-specific detection probabilities can be
%accommodated. Fourth, we presented a Bayesian analysis of the
%model, which, makes it possible to incorporate prior information when
%available, and in some cases, facilitates parameter estimation.

We also demonstrated how many of the objectives of classical
state-space models can be met using this. For example, one of the
primary aims of our paper was to demonstrate how classical
population growth models can be embedded in the DM model.
In some ways, such population growth models are more phenomenological
than mechanistic, and this runs counter to the motivation for the DM
model. However,
%A unique aspect of the Dail and Madsen model as originally
%devised was that it allowed for the estimation of demographic
%parameters under strict distributional assumptions and with the
%assumption of geographic closure. Clearly,
estimating
demographic parameters from count data is an ambitious goal, and
the required assumption will not be valid in many cases, especially
when immigration and emigration occur. This was supported by the
unrealistic estimates of vital rates reported here and by the lack of
support for the mechanistic models relative to the population growth
models.
%One
%advantage of the original DM models over our extensions here is
%that they model population dynamics more mechanistically, by
%focusing on survival and recruitment. As we report here, these
%models often provide unrealistic estimates of these vital rates
%and are generally less supported in model selection.
%Adding
%density dependence directly to one or both vital rates may solve
%both of these problems, although this might require very good
%data.
One strategy for adhereing to the explicit model would be to use
estimates of vital rates from demographic studies as priors in a
Bayesian analysis. Alternatively, count data could be combined with
other data types such as capture-recapture data, which offers a means of
making inferences about demographic processes at broad spatial
scales. Models that are fitted to count data and demographic data are
often referred to as integrated population models
\citep[][IPM]{besbeas_etal:2002, buckland_etal:2004,schaub_etal:2007}
and we believe that a DM model would provide an framework for an
IPM. However, we caution that when count data are affected by movement
and vital rates, IPMs would need to model

@@

Although these integrated
population models may be viewed as the gold standard in state-space
modeling, ecologists are not always so fortunate to have direct
information about vital rates, especially at large spatial scales.
Rather, count data are much more common, and
are produced by many of the largest monitoring programs in the world,
such as the North American Breeding Bird Survey
\citep[BBS;][]{robbins_etal:1986}.

In the context of IPMs, the DM model can be viewed as a specific
manifestation of the general state-space model formulated by
\citep{buckland_etal:2004}. However, their model was designed to
allow for inference about demographic parameters, which will
typically require mark-recapture data. Although this may be viewed
as the gold standard in many studies of wildlife populations, in
this paper, we focus on the analysis of count data, which is the
typical source of information used in state-state modeling
applications, and one of the most widely available forms of
ecological data---it is also the data required by the DM model.

In the absence of direct information about demographic parameters, and
when the original assumptions of the model do not hold, we have
offered extensions of the model to estimate derived parameters such as
population growth rate. This has been the traditional emphasis of
state-space models, and our approach resolves many of the factors
limiting ...

\begin{comment}
  The zero-inflated models for initial abundance and dynamics allow
  one to estimate and model the proportion of sites outside of the
  range of a species.  This could be especially useful when different
  factors control the range than the abundance or dynamics within the
  range.  [some examples where that is the case?]
\end{comment}
[paragraph on random effects]

[paragraph on Bayesian]
The ability to fit this class of models using freely available MCMC
software offers non-statisticians a powerful means .

In spite of the new extensions we have proposed, several aspects of
the model could be improved. First, the precision with which the
parameters of the state process can be estimated ultimately depends
upon how well detection probability is estimated. When there is only a
single survey per primary period, the information about detection
probability comes from deviations from the parametric assumptions
about population dynamics. Thus, without direct information about
detection probability, the estimates will be determined by model based
assumptions. Furthermore, there are multiple components of detection
probability that should be accounted for to minimize bias and yield
valid estimates of population size \citep{nichols_etal:2009}.

Fortunately, it is easy to incorporate direct information about
detection probability and we recommend that this be done whenever
possible. The original method for proposed in the original paper was
to collected replicated counts within the primary periods during a
period in which the populations could be assumed to be closed.  A
robust design could be used to combine multiple surveys per primary
period to increase the precision of the estimates. We envision that
multiple other options are available as well, such as removal, time of
detection, and distance modeling ([cites]).
%For instance, there
%should be no difficulty extending the model to accommodate traditional
%capture-recapture data collected during each primary period. Removal
%or time of detection etc.. Distance-related heterogeneity in detection
%probability is another source of variation that can bias estimators of
%abundance.   Show some equations for the alternative observation
%models.

>>>>

Our work has highlighted the limitations of classical state-space
models as applied to ecological time series data, and we have
demonstrated how a class of open population $N$-mixture models
proposed by \citet{dail_madsen:2011} overcomes these
limitations. We extended their model in several important
ways to accomodate common features of ecological
datasets, including sparse counts, zero-inflation, demographic
stochasticity, and random variation in observation error.
%We made four important developments of the open population
%N-mixture model proposed by \citet{dail_madsen:2011} that make the
%model more applicable to long-term data such as is collected
%by the North American Breeding Bird Survey.
%First, we have
%reconciled the objectives of traditional state-space models
%with open population $N$-mixture models by illustrating how
%classical models of population dynamics can be embedded within
%the framework. Second, we have demonstrated methods of
%accounting for zero-inflation in the time-series. Third, we
%have illustrated how additional random effects such as
%observer-specific detection probabilities can be
%accommodated. Fourth, we presented a Bayesian analysis of the
%model, which, makes it possible to incorporate prior information when
%available, and in some cases, facilitates parameter estimation.

We also demonstrated how many of the objectives of conventional
state-space modeling can be accomplished within this expanded
framework. For example, one of the
primary aims of our paper was to demonstrate how classical
population growth models can be embedded in the DM model. Although we
view this as an important connection to make between the two different
classes of state-space models, we believe that
%In some ways,
such population growth models are more phenomenological
than mechanistic, and this runs counter to the motivation for the DM
model. However, estimating
demographic parameters from count data is an ambitious goal, and
the required assumption will not be valid in many cases, especially
when immigration and emigration occur. This was supported by the
unrealistic estimates of vital rates reported here and by the lack of
support for the mechanistic models relative to the population growth
models.

One strategy for adhereing to the original formulation of the model
that included survival and recruitment parameters would be to use
either prior information or additional data. For example, in a
Bayesian analysis existing estimates of demographic parameters could
be used as informative priors, which might make it possible to
seperate the effects of movement versus births and deaths.
Alternatively, ecologists could combine count data with data from, for example,
capture-recapture data. Models that are fitted to count data and demographic data are
often referred to as integrated population models
\citep[IPM;][]{besbeas_etal:2002, buckland_etal:2004,schaub_etal:2007}
and we believe that a DM model could work well in this context.
In fact, the DM model could be viewed as a specific case of the
IPM model proposed by \citet{buckland_etal:2004}.
However, we caution that when count data are affected by movement
and vital rates, it would be unwise to ignore the movement processes.

@@

Although these integrated
population models may be viewed as the gold standard in state-space
modeling, ecologists are not always so fortunate to have direct
information about vital rates, especially at large spatial scales.
Rather, count data are much more common, and
are produced by many of the largest monitoring programs in the world,
such as the North American Breeding Bird Survey
\citep[BBS;][]{robbins_etal:1986}.
In the absence of direct information about demographic parameters, and
when the original assumptions of the model do not hold, the model can
still be used for important purposes such as predicting future
population size and changes in species distributions.

[paragraph on random effects]

Another important benefit of this class of models is that it can be
readily implemented using frequentist or Bayesian methods by
ecologists lacking skills in computer programming and advanced
statistics. We have included an appendix with extensive code to
simulate and fit many of the extensions discussed in this paper.

In spite of the new extensions we have proposed, several aspects of
the model could be improved. First, the precision with which the
parameters of the state process can be estimated ultimately depends
upon how well detection probability is estimated. When there is only a
single survey per primary period, the information about detection
probability comes from deviations from the parametric assumptions
about population dynamics. Thus, without direct information about
detection probability, the estimates will be determined by the model's
parametric assumptions. Furthermore, properly modeling detection
probability can involve account for nuances such as difference among
individuals and probabilities of being available for detection
\citep{nichols_etal:2009}.

Fortunately, it is straight-forward to incorporate direct information about
detection probability and we recommend that this be done whenever
possible. The original method for proposed in the original paper was
to collected replicated counts within the primary periods during a
period in which the populations could be assumed to be closed.  A
robust design could be used to combine multiple surveys per primary
period to increase the precision of the estimates. We envision that
multiple other options are available as well, such as removal, time of
detection, and distance modeling ([cites]). These could be
accomplished by extending the open population $N$-mixture model in
exactly the same way as the closed population version has been
extended \citep[e.g.,][]{royle_etal:2004}.
